Reproducible methods for modeling combined public transport and cycling trips and associated benefits: evidence from the biclaR tool

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.tex
beef spatial results and re-structure conclusions

@@

The Lisbon metropolitan area (LMA), in Portugal, is the case study,
comprising 18 municipalities, with a total area of 3,015 km² and a
population of 2.8 million residents The LMA is the most populous and
economically active region in Portugal, with the city of Lisbon as the
central node of the LMA, with a high concentration of jobs and services,
and a high demand for mobility (see Figure \ref{fig:lma}). As shown, the
number of trips is proportional to the population of each municipality,
with the highest number of trips originating in Lisbon, followed by the
municipalities of Sintra, Cascais, Loures, and Almada.

The Public Transport network is well developed, although very much
Lisbon-centric, with a high level of service, including inter-municipal
trains, buses, trams, and ferries (see Figure \ref{fig:map1}), and a
metro system that serves Lisbon and surrounding municipalities.

\begin{figure}
\subfloat[Population in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA according to census and the travel survey.}\label{fig:lma}
\end{figure}

@@

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs. We restricted our analysis to the
first and last legs with a combined length of up to 5 km (for example: 1
km from origin to interface \emph{A} plus 4 km from interface \emph{B}
to destination). Furthermore, we have imposed restrictions on PT usage,
limiting it to trips without PT transfers, and within a duration of up
to 2 hours (120 minutes). Additionally, we have only included PT modes
that can easily accommodate bicycles, such as trains, ferries, trams,
and inter-municipal bus lines equipped with bike racks (Figure
\ref{fig:map1}).

@@

Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
the main PT interfaces in the LMA. It is noticeable the high potential
of the train interfaces to attract car-to-PT substituting trips.

@@

\subsection{Modelling car shift to bicycle-PT - here or
before??}\label{modelling-car-shift-to-bicycle-pt---here-or-before}

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
daily trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment (SE) benefits are presented in Table
\ref{tab:summary1b}, resulting from replacing car trips with cycling.

@@

The benefits are spatially unevenly distributed across the LMA, as shown
in Figure \ref{fig:mapzones}.

Table \ref{tab:summary22} presents emissions reductions and associated
economic benefits associated with the second (PT) leg of trips. The
shift from private car associated with thes PT segments would reduce
CO\textsubscript{2} equivalent emissions by 8,500 to 20,800 tons
annually, valued in €1.4 million to €3.5 million yearly, for the 4\% and
10\% targets, respectively.

@@

Our findings show that cycling \emph{in combination} with PT could
replace 10\% of current LMA trips, with an additional 6\% of PT journeys
prone to further substitution, based on conservative assumptions.
Although this paper does not address the designing specific solutions,
our findings indicate a potential surge in demand (+6.3\%) of public
transport interfaces to align with the national strategies. This added
demand is conceptually driven from individuals utilising bicycles,
either personal or shared, which would be parked at the PT stations.
Consequently, operators and public space managers would face the
challenge of accommodating this increased demand for bicycle parking
and/or docks from a bike sharing system, since this availability plays a
crucial role in the decision-making process of individuals to use
combined bike-train \citep{jonkeren2021bicycle}. Additionally, they may
need to explore options to enhance bicycle carriage capacity on public
transport services, ensuring alignment with national targets.

@@

This paper estimated the potential for combining cycling and PT to
substitute car trips in the LMA, while achieving the national cycling
targets and supporting decarbonization goals. It has become
progressively more common to establish strategic plans, at national,
regional or municipal level, to mitigate climate change. Among these,
the Sustainable Urban Mobility Plans (SUMP\footnote{See
  \href{https://www.eltis.org/mobility-plans/sump-concept}{eltis.org/mobility-plans/sump-concept}}),
promoted by the European Commission, are becoming popular in Europe,
although authorities are designing documents of this sort all over the
world. The definition of targets associated with a timeframe for
reducing dependence on the individual motorized vehicle, or targets for
the use of active modes such as walking and cycling, are too often not
accompanied by estimates of their social, environmental and economic
impacts. It is important for authorities and practitioners to know how
to estimate those impacts, which tools are available to support them in
the process, and what results to expect.

This paper quantifies the benefits of replacing car trips with cycling,
in combination with public transport. The case study of the Lisbon
metropolitan area demonstrates that cycling-PT integration can help meet
the national targets set for bicycle use of 4\% and 10\% by 2025 and
2030, respectively.

The quantification of benefits can support policy-makers in prioritizing
interventions to reduce the reliance on private motorized modes of
transportation. The presentation of the results in an open access web
application will help to inform and explain decisions. Furthermore, the
provision of datasets resulting from this project provides a foundation
for further research and development of new tools and methods. The
methods are reproducible and based on open source software, which can be
applied to other cities and metropolitan areas, supporting the
decarbonization of transport systems internationally.

>>>>

It has become progressively more common to establish strategic plans, at
national, regional or municipal level, to mitigate climate change. Among
these, the Sustainable Urban Mobility Plans (SUMP\footnote{See
  \href{https://www.eltis.org/mobility-plans/sump-concept}{eltis.org/mobility-plans/sump-concept}}),
promoted by the European Commission, are becoming popular in Europe,
although authorities are designing documents of this sort all over the
world. The definition of targets associated with a timeframe for
reducing dependence on the individual motorized vehicle, or targets for
the use of active modes such as walking and cycling, are too often not
accompanied by estimates of their social, environmental and economic
impacts. It is important for authorities and practitioners to know how
to estimate those impacts, which tools are available to support them in
the process, and what results to expect.

@@

The Lisbon metropolitan area (LMA), in Portugal, is the case study,
comprising 18 municipalities, with a total area of 3,015 km² and a
population of 2.8 million residents The LMA is the most populous and
economically active region in Portugal, with the city of Lisbon as the
central node of the LMA, with a high concentration of jobs and services,
and a high demand for mobility (see Figure \ref{fig:lma}).

As shown, the number of trips is proportional to the population of each
municipality, with the highest number of trips originating in Lisbon,
followed by the municipalities of Sintra, Cascais, Loures, and Almada.
The Public Transport network is well developed, although very much
Lisbon-centric, with a high level of service, including inter-municipal
trains, buses, trams, and ferries (see Figure \ref{fig:map1}), and a
metro system that serves Lisbon and surrounding municipalities.

\begin{figure}
\subfloat[Population distribution in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to census and the travel survey.}\label{fig:lma}
\end{figure}

@@

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs. We restricted our analysis to the
first and last legs with a combined length of up to 5 km (for example: 1
km from origin to interface \emph{A} plus 4 km from interface \emph{B}
to destination). Furthermore, we have imposed restrictions on PT usage,
limiting it to trips without PT transfers, and within a duration of up
to 2 hours (120 minutes). Additionally, we have only included PT modes
that can easily accommodate bicycles, such as trains, ferries, trams,
and inter-municipal bus lines equipped with bike racks (Figure
\ref{fig:map1}). This map shows that some of the municipalities are not
served by PT modes that can easily accommodate bicycles, which may limit
the potential for intermodality in these areas.

@@

Figure \ref{fig:map2} illustrates the routes with greatest cycling
potential to access nodes on the public transport network. The
interactive version of the map is publicly available in the
\emph{biclaR} tool, allowing detailed exploration of local areas. At
\emph{biclaR}, users and local authority practitioners have the option
to download the results as a spatial dataset for further analysis
concerning the prioritization of cycling interventions. It is noticeable
the high potential of the train interfaces to attract car-to-PT
substituting trips.

@@

\subsection{Modelling car shift to
bicycle-PT}\label{modelling-car-shift-to-bicycle-pt}

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
daily trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles, at an aggregate level of the LMA. For the
cycling legs of the journey (first and last legs), the environmental
avoided emissions and monetized socio-environment (SE) benefits are
presented in Table \ref{tab:summary1b}, resulting from replacing car
trips with cycling.

@@

Another relevant aspect is that the higher potential (Figure
\ref{fig:mapzones}.a) is in the municipalities with the highest number
of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and
Almada, which are also the ones with the highest number of PT interfaces
(Figure \ref{fig:map1}). On another hand, the municipalities with little
PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a
lower potential for car to Bike-PT substitution (Figure
\ref{fig:mapzones}.a), even if they have a high number of trips, such as
Setubal. This findings are aligned with their local travel patterns,
where the majority of trips are intra-municipal (67\% for both
municipalities), and not to the city of Lisbon, despite the high
car-dependency (79\% in Mafra, 67\% in Setubal).

The potential and the resulting benefits are spatially unevenly
distributed across the LMA, as shown in Figure \ref{fig:mapzones}.b,
concentrating 40\% in Lisbon, where the highest number of trips are
made.

\begin{figure}
\subfloat[Potential new Bicycle + PT trips\label{fig:mapzones-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-1} }\subfloat[Proportion of total potential modal shift in the LMA\label{fig:mapzones-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-2} }\caption{\label{mapzones}Potential Bicycle + PT daily trips for the 10\% target and direct routing, by municipality. The percentages are shown for municipalities with values above 4\%.}\label{fig:mapzones}
\end{figure}

Table \ref{tab:summary22} presents emissions reductions and associated
economic benefits associated with the second (PT) leg of trips. The
shift from private car associated with the PT segments would reduce
CO\textsubscript{2} equivalent emissions by 8,500 to 20,800 tons
annually, valued in €1.4 million to €3.5 million yearly, for the 4\% and
10\% targets, respectively.

@@

Our findings show that cycling \emph{in combination} with PT could
replace 10\% of current LMA trips, with an additional 6\% of PT journeys
prone to further substitution, based on conservative assumptions,
highlighting the municipalities of Lisbon, Sintra, Almada, Amadora, and
Oeiras as the ones with the highest potential for car to bike-PT
substitution. Although this paper does not address the designing
specific solutions, our findings indicate a potential surge in demand
(+6.3\%) of public transport interfaces to align with the national
strategies. This added demand is conceptually driven from individuals
utilising bicycles, either personal or shared, which would be parked at
the PT stations. Consequently, operators and public space managers would
face the challenge of accommodating this increased demand for bicycle
parking and/or docks from a bike sharing system, since this availability
plays a crucial role in the decision-making process of individuals to
use combined bike-train \citep{jonkeren2021bicycle}. Additionally, they
may need to explore options to enhance bicycle carriage capacity on
public transport services, ensuring alignment with national targets.

@@

This paper estimates the potential for combining cycling and PT to
substitute car trips in the LMA, while achieving the national cycling
targets and supporting decarbonization goals. The case study of the
Lisbon metropolitan area demonstrates that cycling-PT integration can
help meet the national targets set for bicycle use of 4\% and 10\% by
2025 and 2030, respectively.

This research also quantifies the socio-environmental benefits of
achieving such targets, exploring an intermodality scenario, where car
trips are potentially substituted by bicycle in combination with PT.

The findings indicate that cycling combined with public transport has
the potential to account for 10\% of total trips. Transitioning to
cycling for the first and last segments could lead to a decrease in
annual CO\textsubscript{2}eq emissions by a substantial margin, ranging
from 3,000 to 7,500 tons per day. Moreover, for the second trip segment,
the shift from car to public transport contributes to avoiding up to
20,500 tons of CO\textsubscript{2}eq emissions annually. These changes
are estimated to yield socio-environmental benefits totaling €125
million to €325 million over a decade. The quantification of such
benefits can support policy-makers in prioritizing interventions to
reduce the reliance on private motorized modes of transportation.

Additionally, the results suggest that the potential for cycling-PT
intermodality is spatially unevenly distributed across the LMA, with the
highest potential in the municipalities with the highest number of trips
and PT interfaces of PT modes connecting to Lisbon and where it is
possible to carry bicycles on board, such as Lisbon, Sintra, Amadora,
and Almada.

We opted to detail all the steps of the methodology, from data
requirements to the final results, to ensure transparency and
generalization of the approach. We acknowledge the limitations of the
used methods and propose avenues for future research.

The presentation of the results in an open access web application will
help to inform and explain decisions. Furthermore, the provision of
datasets resulting from this project provides a foundation for further
research and development of new tools and methods. The methods are
reproducible and based on open source software, which can be applied to
other cities and metropolitan areas, supporting the decarbonization of
transport systems internationally.
